Chemical Tagging Using Minimum Message Length. I.\\ Gaussian Mixture Modelling

### 2017-05-01 articles/chemical-tagging-gmm/ms.tex
Expand math for model, some minor text changes

\begin{abstract}
Chemical tagging seeks to identify co-natal stars by their present-day 
photospheric abundances, long after their phase space similarity is lost.
In principle the detailed chemical abundances of a transformative number of
stars could be used to reconstruct the evolution of the Milky Way, providing
insight on supernovae yields, galactic dynamics (e.g., constraints on radial 
mixing), the initial mass function, as well as the formation and destruction
of star clusters. Some progress has been made towards chemical tagging, but 
here we argue that even the \emph{extent} of the problem has not been realised.
We introduce the Real World problems associated with chemical tagging at scale,
and describe how currently considered methods will inevitably fail even under
the most unrealistically optimistic conditions. We introduce the minimum message
length to the astronomical community as a promising alternative to chemical
tagging, one which has the capability to address all ensivaged problems with
chemical tagging simultaneously. We perform chemical tagging experiments with
infinte Gaussian mixture models, and demonstrate using real and generated data 
that MML outperforms all other considered penalty techniques. 
\end{abstract}

@@

\citet{Freeman_2002} introduced the idea of chemical tagging to identify 
groups of stars that formed together in the same gas cloud, which are no 
longer identifiable from their physical proximity or orbital properties.
This idea is attractive because the observable chemical abundances of 
stars remains largely unchanged throughout a star's lifetime, whereas phase 
space similarity can be quickly washed out through dynamical interactions. 
Indeed, although most stars are thought to form in clusters \citep{someone},
only the most massive or isolated star clusters are able to survive the star
formation process \citep{someone2}, leaving most clusters to become 
dynamically unbound after about 1~Gyr.

@@

%The Milky Way is our best laboratory for understanding galaxy formation and
%evolution. 
Making strong inferences about the Milky Way's formation and evolution from
chemical tagging will require a transformative number of stars with precise
abundances.
Ground- and space-based surveys are poised to deliver those data this decade. 
The Apache Point Observatory Galactic Evolution Experiment 
\citep[\apogee, e.g.,][]{apogee} and the Gaia-ESO survey 
\citep[\ges;][]{Gilmore_2012,Randich_2013} are already delivering $10^5$ stars 
with 20--40 precise abundances, and the Galactic Archaeology with \hermes\ 
\citep[\galah;][]{DeSilva_2015} survey seeks to derive $\approx$30 precise 
abundances for $\sim10^6$ stars. 
Future ground-based surveys will derive abundances from $\approx$20 million 
spectra in each hemisphere \citep[e.g., \weave, \fourmost;][]{weave,4most},
and the radial velocity spectrometer on \gaia\ can deliver a few abundances 
for $\approx$150 million stars.

Chemical tagging is a simple idea, but with this volume of data it represents
a formidable parameter estimation and model selection problem.
Before reviewing the literature on chemical tagging, here we explicitly
formulate the scope of the analysis problem.

\subsection{What do we want to infer from chemical abundances?}
\label{sec:the-problem}

@@

\begin{enumerate}
  \item The number of star clusters (or star-forming events) where at least 
        one star was observed and chemical abundances were derived.
  \item The conditional probabilities (or memberships) of each star belonging
        to every star cluster or star-forming event.
        Alternatively, the relative weighting of individual mixtures.
  \item The mean abundances of each star cluster. 
  \item The covariance matrix of abundances of each star cluster.
        This covariance matrix includes the intrinsic (cluster) variance
        of individual abundances (e.g., the homogeniety), and allows for
        correlations between abundances.\footnote{
          It is likely -- or hoped -- that \emph{true} correlations between 
          abundances will be accounted for through the latent factors, and any 
          remaining correlations are the result of the measurements.}
  \item The \emph{number} of common multivariate latent factors, variables 
        that are not observable but affect all or many of the observations.
        In the context of chemical tagging, the yields of core-collapse
        supernovae can be considered a common multivariate latent factor: 
        those yields are not directly observable, but they contribute to much 
        of the data.
        Here we are interested in the \emph{number} of common multivariate
        latent factors. For example, the number of sources of enrichment:
        different classes of supernovae, asymptotic giant branch (AGB) stars, 
        etc.
  \item The multivariate factor loads of individual latent factors. 
        For example, if core-collapse supernovae and AGB stars had both 
        contributed to the chemical abundances of all stars in the sample, 
        then the \emph{factor loads} would be the \emph{yield} of the 
        supernovae, or the yield of relative abundances that were produced by 
        the AGB star. 
        Each star will be affected differently by those yields (see item 7).
  \item The relative \emph{scoring} of different latent factors on each star.
        This is a scaling that is applied to the latent yields. Continuing our
        analogy on yields from core-collapse supernovae and AGB stars, the
        \emph{factor scores} can be thought of as the relative contributions 
        of the different factor loads, for a single star.
  \item The intrinsic \emph{specific variances} of the individual abundances.
        These variances can be thought of as the intrinsic variability from 
        the different factor loads.
        The intrisic variances affect all observations, before accounting for
        any homogeneity in individual clusters.
  \item The \emph{number} of stellar age- or parameter-dependent latent 
        factors that affect the data. 
        The surface chemical abundances of stars \emph{do change} over its
        lifetime, at least in part due to atomic diffusion, veiling, and 
        thermohaling mixing.
        Understanding and modelling these effects will be critical for 
        chemical tagging stars across all stellar ages and evolutionary states.
  \item The factor loads of stellar age- or parameter-dependent latent factors.
        In this example, a single multivariate factor load may represent the
        age-dependent effects of atomic diffusion.
  \item The factor scores of stellar age- or parameter-dependent latent 
        factors.
\end{enumerate}

@@

\subsection{Chemical tagging progress to date}
\label{sec:literature-review}

There has been substantial progress on chemical tagging in the last decade.
Given the difficulty of the problem, most studies have opted to address one
(or a few) of the challenges listed in Section~\ref{sec:the-problem}. 
The most common issue addressed is group-finding (items 1-4). 
Some works have sought to construct a similarity metric \citep{Mitschang}
based on chemical abundances of undisrupted clusters. 
Others have applied group-finding algorithms \citep{Hogg,Gregor}.

% TODO
- Dimensionality of chemical abundance space.

- Chemical homogeneity.

- Dopplegangers.

The aforementioned chemical tagging experiments have sought to address only 
individual challenges associated with chemical tagging.
These challenges are all related, and therefore addressing them separately
will yield results that are not self-consistent, and do not make use of all
the information available.

@@

\noindent{}The probability density function $p$ for a multivariate normal 
distribution with $D$ dimensions is given by

\begin{equation}
  p(\data|\vecmean,\veccov) 
      = \frac{1}{\sqrt{(2\pi)^d|\veccov|}}
        \exp{\left[-\frac{1}{2}(\data - \vecmean)^\intercal\veccov^{-1}(\data - \vecmean)\right]}
\end{equation}

@@

\begin{equation}
  p(\data|\vectheta) = \sum_{k=1}^{K} \weight_k \, p(\data|\vectheta_k)
\end{equation}

\noindent{}where $\vectheta_k \equiv \{\vecmean_k, \veccov_k\}$,
$\vectheta \equiv \{\vectheta_1,\dots,\vectheta_K,\weight_1,\dots,\weight_K\}$, and
$\weight_k$ is the relative weight (or the mixing probability) of the $k$-th 
mixture,\footnote{
  An alternative representation would be to include integer variables for each
  of the $N$ stars, where each variable represents the cluster which that star
  belongs.
  These two models are mathematically identical \citep{Foreman_Mackey_2014},
  but the integer representation is highly dimensional and difficult to
  optimize.}
and we require that $\sum_{k=1}^{K} \weight_k = 1$.
Although the true number of $K$ mixtures is unknown (as are the means $\vecmean$
and covariance matrices $\veccov$), we will address the selection of $K$
and parameter estimates of $\vectheta$ in Section \ref{sec:fitting-the-data}.

@@

\noindent{}which states that the joint probability of the data $D$ and 
hypothesis $H$, $P(H, D)$, is given by

\begin{equation}
  P(H, D) = P(H) \cdot P(D|H) = P(D) \cdot P(H|D)
\end{equation}

@@

\begin{equation}
  I(H, D) = I(H) + I(D|H) = I(D) + I(H|D) \quad .
  \label{eq:mml-wallace}
\end{equation}

@@

Calculating the length of the message is a non-trivial task, especially if the
model is even reasonably complex.
This makes the MML principle intractable (or uncomputable) in most cases, and
forces us to make approximations when calculating the message length.
\citet{Wallace_Freeman_1987} introduce a generalised scheme to estimate a
vector of parameters $\vectheta$ of any distribution given some data, where
the estimates $\vectheta$ that minimise the message length expression,

\begin{equation}
  I(\vectheta,\data) 
    = \frac{p}{2}\log{q_p} - \log{\left(\frac{h(\vectheta)}{\sqrt{|\mathcal{F}(\vectheta)|}}\right)}
      - \likelihood\left(\data|\vectheta\right) + \frac{p}{2}
      \label{eq:message-length}
\end{equation}

\noindent{}is the MML estimate.
Here $h(\vectheta)$ is the prior on the hypothesis, $p$ is the number of free
parameters in the model, $q_p$ is the lattice quantization constant
\citep[e.g.,][]{Conway_1984}, and $|\mathcal{F}\left(\vectheta\right)|$ is the
determinant of the \emph{expected} Fisher information matrix (the second-order
partial derivatives of the negative log-likelihood function 
$-\likelihood(\data|\vectheta)$).

@@

The MML principle requires that the prior beliefs of the model to be fully 
specified as part of the message. 
We must explicitly specify our prior beliefs on the model parameters described
in Section \ref{sec:model} in order to minimize the full message length.
Indeed, the message must fully specify all components necessary for the
receiver to be able to reconstruct the message.
For the model specified in Section \ref{sec:model}, this requirement forces us
to specify a number of components:

\begin{enumerate}
  \item The number of Gaussian mixtures, $K$.
  \item Encoding the relative weights $\weights$ of the Gaussian mixtures.
  \item Encoding the component parameters $\vecmean$, $\veccov$ for all of
        the Gaussian mixtures.
  \item Encoding the data, given the model parameters.
  \item The lattice quantization constant $q_p$ for the number of model
        parameters $p$.
\end{enumerate}

Some of these components require non-trivial derivations or approximations in
order to calculate the length of the message required to encode the component.
For this reason, we will consider the message length of each component in turn
before returning to the full message length specified in Equation \ref{eq:message-length}.

@@

We assume that fewer mixtures are more likely than a larger number of mixtures.
Specifically, we assume that $p(K) \propto 2^{-K}$, which implies that the
message length required to encode $K$ is given by:

@@

We assume a uniform prior on the mixing weights $\bm{w}$, only requiring that
$\sum_{k=1}^{K}w_k = 1$.
% This implies the weights are parameters of a multinomial distribution.
\citet{Boulton_1969} give the following expression for the message length 
required to encode the weights,

\begin{equation}
I(\weights) = \frac{K - 1}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\weight_k - \log{\Gamma{\left(K\right)}} \quad ,
\end{equation}

@@

Recall that what we need to encode is both our priors on the model parameters
and the determinant of the expected fisher information matrix.
For the $k$-th mixture this becomes,

@@

\begin{equation}
I(\vecmean,\veccov) = -\sum_{k=1}^{K}\log{p(\vecmean_k,\veccov_k)} + \frac{1}{2}\sum_{k=1}^{K}\log{|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|}
\end{equation}

This requires us to explicitly specify our priors on the parameters $\vecmean$
and $\veccov$.
We adopt an improper uniform prior of 
$\mathcal{U}(\vecmean) = [-\infinity, \infinity]$ on all multivariate mean 
abundances $\vecmean$ (in all $D$ dimensions).
This prior has no impact on our inferences and becomes proper when it enters 
the (accepted) posterior distribution. 
We similarily could adopt a proper and large uniform prior on all mean 
abundances of $\mathcal{U}(\vecmean) = [-12, 12]$ (24 orders of magnitude!) 
and the result would be the same.

We adopt a congujate inverted Wishart prior for the covariance matrices in the
individual mixtures. 
The joint prior density for the parameters of a single
mixture is given by the limiting form of the normal inverted-Wishart
density \citep[e.g., Section 5.2.3 of ][]{Schafer_1990}:

\begin{equation}
    p(\vecmean_k,\veccov_k) \propto |\veccov_k|^{\frac{D + 1}{2}} \quad .
\end{equation}

% The fisher information matrix

\subsection{Prior probabilities on the parameters}
\label{sec:priors}

\subsection{Expected Fisher information}

Computing the expected Fisher information requires the second order partial
derivatives of the negative log-likelihood $-\likelihood(\data|\vecmean,\veccov)$.
This is difficult to calculate for mixtures of Gaussians, and for this reason
we approximate the determinant of the Fisher information 
$|\mathcal{F}(\vecmean, \veccov)|$ as the product of 
$|\mathcal{F}\left(\vecmean\right)|$ and $|\mathcal{F}\left(\veccov\right)|$ 
\citep{Oliver_1996,Roberts_1998}. 
We begin by taking second derivative of the log-likelihood function
$\likelihood\left(\data|\vecmean,\veccov\right)$,

@@

\noindent{}and for a $d$-dimensional vector $\vecmean$ the Fisher information
matrix is:

\begin{eqnarray}
  |\mathcal{F}(\vecmean)| = \left|-\frac{\delta^2\likelihood}{\delta\vecmean^2}\right| = N^d|\veccov|^{-1} \quad .
\end{eqnarray}

Computing the Fisher information of the covariance matrix 
$|\mathcal{F}\left(\veccov\right)|$ is harder. 
For a full covariance matrix with off-diagonal terms, the total number of free 
parameters required is $d(d+1)/2$.
\citep{Magnus_1988} derived an analytical expression:

% Dwyer_1967, Bozdogan_1990? Drton_2009?
% TODO: This derivation is a nightmare to find

\begin{equation}
  |\mathcal{F}\left(\veccov\right)| = N^\frac{d(d+1)}{2}2^{-d}|\veccov|^{-(d+1)}
\end{equation}

\noindent{}Giving the approximate determinant of the expected Fisher information 
$|\mathcal{F}\left(\vecmean,\veccov\right)|$ as:

\begin{eqnarray}
  |\mathcal{F}\left(\vecmean,\veccov\right)| & \approx & |\mathcal{F}\left(\vecmean\right)|\cdot|\mathcal{F}\left(\veccov\right)| \nonumber \\
  |\mathcal{F}\left(\vecmean,\veccov\right)| & \approx & N^{d}|\veccov|^{-1}N^\frac{d(d+1)}{2}2^{-d}|\veccov|^{-(d+1)} \nonumber \\
  |\mathcal{F}\left(\vecmean,\veccov\right)| & \approx & N^\frac{d(d+3)}{2}2^{-d}|\veccov|^{-(d+2)} \quad .
\end{eqnarray}

\subsection{Encoding the message}

Given our priors, likelihood function, and approximate expressions for the
Fisher information, we are now in a position to calculate the total message
length.
There are a number of components that go into the message:

A common method to estimate the parameters of a Gaussian distribution is by
maximum likelihood, where the log-likelihood $\likelihood$ is maximized. In
order to do this, one computes the gradient of the log-likelihood function
with respect to the parameters and solves the resulting equations.

- Encoding the message.

- Process of expectation and maximization for each step.

>>>>

\begin{abstract}
Chemical tagging seeks to identify co-natal stars by their present-day 
photospheric abundances, long after their phase space similarity is lost.
In principle the detailed chemical abundances of a transformative number of
stars could be used to reconstruct the evolution of the Milky Way, providing
insight on supernovae yields, galactic dynamics, the initial mass function, 
as well as the formation and destruction of star clusters.

Some progress has been made towards chemical tagging, but 
here we argue that even the \emph{extent} of the problem has not been realised.
We introduce the Real World problems associated with chemical tagging at scale,
and describe how currently considered methods will inevitably fail even under
the most unrealistically optimistic conditions. We introduce the minimum message
length to the astronomical community as a promising alternative to chemical
tagging, one which has the capability to address all ensivaged problems with
chemical tagging simultaneously. We perform chemical tagging experiments with
infinte Gaussian mixture models, and demonstrate using real and generated data 
that MML outperforms all other considered penalty techniques. 
\end{abstract}

@@

\citet{Freeman_2002} introduced the idea of chemical tagging to identify 
groups of stars that formed together in the same gas cloud, which are no 
longer identifiable from their physical proximity or orbital properties.
This idea is attractive because the observable chemical abundances of 
stars remains largely unchanged throughout a star's lifetime, whereas phase 
space similarity can be quickly washed out through dynamical interactions. 
Indeed, although most stars are thought to form in clusters \citep{someone},
only the most massive or isolated star clusters are able to survive the star
formation process \citep[e.g.,][]{Lada_2010}, leaving most clusters to become 
dynamically unbound after about 1~Gyr.

@@

Despite the Milky Way being our best laboratory for understanding galaxy
formation, a transformative number of stars with precise abundances will be
required to chemically tag the Galaxy.
Ground- and space-based surveys are poised to deliver those data in the coming
decade. 
The Apache Point Observatory Galactic Evolution Experiment 
\citep[\apogee, e.g.,][]{apogee} and the Gaia-ESO survey 
\citep[\ges;][]{Gilmore_2012,Randich_2013} are already delivering $10^5$ stars 
with 20--40 precise abundances, and the Galactic Archaeology with \hermes\ 
\citep[\galah;][]{DeSilva_2015} survey seeks to derive $\approx$30 precise 
abundances for $\sim10^6$ stars. 
Future ground-based surveys will derive abundances from $\approx$20 million 
spectra in each hemisphere \citep[e.g., \weave, \fourmost;][]{weave,4most},
and the radial velocity spectrometer on \gaia\ can deliver a few abundances 
for $\approx$150 million stars.

Chemical tagging is a simple idea, but with this volume of data it represents
a formidable parameter estimation and model selection problem.
Given the difficulty of the problem, most chemical tagging studies have 
focussed on addressing one \emph{aspect} of chemical tagging.
For example, even the simplest applications of chemical tagging requires the
identification of groups in the data, with stars being assigned to certain
groups (a clustering problem), and to somehow chose the number of groups that
are best represented by the data (a model selection problem).
Previous works include definitions of L1-like metrics \citep{Mitschang_2014},
where open clusters are used to define confidence intervals of similarity
\citep{Mitschang_2013}, and whether similar stars constitute a group or not.
Other studies have directly applied standard clustering algorithms,
\citep[e.g., \acronym{DBSCAN}, $k$-means;][]{Blanco_Cuaresma_2015,Hogg_2016} 
and fixed the number of clusters.
Inferring the number of true clusters -- one of the principal goals of 
chemical tagging -- was not attempted in these simplified experiments.

Other works have focussed [TODO EXPAND]:
\begin{itemize}
  \item Dimensionality of chemical abundance space (Ting et al).
  \item Chemical homogenity of open clusters (Bovy, Da Silva)
  \item Dopplegangers (Ness)
  \item A single common latent factor, integrated over cosmic time (Bland-Hawthorn)
\end{itemize}

The aforementioned chemical tagging experiments have sought to address only 
individual challenges associated with chemical tagging.
These challenges are all related, and therefore addressing them separately
will give results that ado not make use of all
the information available. 
Here, as we introduce the problem of chemical tagging, we list the inferences
we wish to make from a large number of chemical abundances.

\subsection{What's the goal of chemical tagging? What do we want to infer?}
\label{sec:the-problem}

@@

\begin{enumerate}
  \item The number of star clusters (or star-forming events) where at least 
        one star was observed and chemical abundances were derived.
  \item The conditional probabilities (or memberships) of each star belonging
        to every star cluster or star-forming event.
        Alternatively, the relative weighting of individual mixtures.
  \item The mean abundances of each star cluster. 
  \item The covariance matrix of abundances of each star cluster.
        This covariance matrix includes the intrinsic (cluster) variance
        of individual abundances (e.g., the homogeniety), and allows for
        correlations between abundances.\footnote{
          It is likely -- or hoped -- that \emph{true} correlations between 
          abundances will be accounted for through the latent factors, and any 
          remaining correlations are the result of the measurements.}
  \item The \emph{number} of common multivariate latent factors, variables 
        that are not observable but affect all or many of the observations.
        In the context of chemical tagging, the yields of core-collapse
        supernovae can be considered a common multivariate latent factor: 
        those yields are not directly observable, but they contribute to much 
        of the data.
        Here we are interested in the \emph{number} of common multivariate
        latent factors. For example, the number of sources of enrichment:
        different classes of supernovae, asymptotic giant branch (AGB) stars, 
        etc.
  \item The multivariate factor loads of individual latent factors. 
        For example, if core-collapse supernovae and AGB stars had both 
        contributed to the chemical abundances of all stars in the sample, 
        then the \emph{factor loads} would be the \emph{yield} of the 
        supernovae, or the yield of relative abundances that were produced by 
        the AGB star. 
        Each star will be affected differently by those yields (see item 7).
  \item The relative \emph{scoring} of different latent factors on each star.
        This is a scaling that is applied to the latent yields. Continuing our
        analogy on yields from core-collapse supernovae and AGB stars, the
        \emph{factor scores} can be thought of as the relative contributions 
        of the different factor loads, for a single star.
  \item The intrinsic \emph{specific variances} of the individual abundances.
        These variances can be thought of as the intrinsic variability from 
        the different factor loads.
        The intrisic variances affect all observations, before accounting for
        any homogeneity in individual clusters.
  \item The \emph{number} of stellar age- or parameter-dependent latent 
        factors that affect the data. 
        The surface chemical abundances of stars \emph{do change} over its
        lifetime, at least in part due to atomic diffusion, veiling, and 
        thermohaling mixing. % TODO: Add Korn, others?
        Understanding and modelling these effects will be critical for 
        chemical tagging stars across all stellar ages and evolutionary states.
  \item The factor loads of stellar age- or parameter-dependent latent factors.
        In this example, a single multivariate factor load may represent the
        age-dependent effects of atomic diffusion.
  \item The factor scores of stellar age- or parameter-dependent latent 
        factors.
\end{enumerate}

@@

@@

\noindent{}The probability density function $f$ for a multivariate normal 
distribution with $D$ dimensions is given by

\begin{equation}
  f(\data|\vecmean,\veccov) 
      = \frac{1}{\sqrt{(2\pi)^d|\veccov|}}
        \exp{\left[-\frac{1}{2}(\data - \vecmean)^\intercal\veccov^{-1}(\data - \vecmean)\right]}
\end{equation}

@@

\begin{equation}
  f(\data|\vectheta) = \sum_{k=1}^{K} \weight_k \, f(\data|\vectheta_k)
\end{equation}

\noindent{}where $\vectheta_k \equiv \{\vecmean_k, \veccov_k\}$,
$\vectheta \equiv \{\vectheta_1,\dots,\vectheta_K,\weight_1,\dots,\weight_K\}$, and
$\weight_k$ is the relative weight (or the mixing probability) of the $k$-th 
mixture,\footnote{
  An alternative representation would be to include integer variables for each
  of the $N$ stars, where each variable represents the cluster which that star
  belongs.
  These two models are mathematically identical \citep[e.g., see discussion in][]{Foreman_Mackey_2014},
  but the integer representation is highly dimensional and difficult to
  optimize.}
and we require that $\sum_{k=1}^{K} \weight_k = 1$.
Although the true number of $K$ mixtures is unknown (as are the means $\vecmean$
and covariance matrices $\veccov$), we will address the selection of $K$
and parameter estimates of $\vectheta$ in Section \ref{sec:fitting-the-data}.

@@

\noindent{}which states that the joint probability of the data $D$ and 
hypothesis $H$, $P(H,D)$, is given by

\begin{equation}
  P(H,D) = P(H) \cdot P(D|H) = P(D) \cdot P(H|D)
\end{equation}

@@

\begin{equation}
  I(H,D) = I(H) + I(D|H) = I(D) + I(H|D) \quad .
  \label{eq:mml-wallace}
\end{equation}

@@

Calculating the length of the message is a non-trivial task, especially if the
model is even reasonably complex.
This makes the MML principle intractable (or uncomputable) in most cases, and
forces us to make approximations when calculating the message length.
Using a Taylor expansion, \citet{Wallace_1987} introduce a generalised 
scheme to estimate a parameter vector $\vectheta$ (of any distribution), that
minimises the message length expression:
\begin{equation}
\label{eq:message-length}
  I(\vectheta,\data) 
    = \frac{Q}{2}\log\kappa{\left(Q\right)} 
    - \log{\left(\frac{p(\vectheta)}{\sqrt{|\mathcal{F}(\vectheta)|}}\right)}
    - \likelihood\left(\data|\vectheta\right) 
    + \frac{Q}{2} \quad .
\end{equation}

Here $p(\vectheta)$ is the prior on the hypothesis, $q$ is the number of free
parameters in the model, $\kappa\left(Q\right)$ is a function to approximate the lattice
quantisation constant for $Q$ free parameters \citep[e.g.,][]{Conway_1984}, 
and $|\mathcal{F}\left(\vectheta\right)|$ is the determinant of the 
\emph{expected} Fisher information matrix (the second-order partial 
derivatives of the negative log-likelihood function 
$-\likelihood(\data|\vectheta)$).

@@

The MML principle requires that the message fully specifies all components
necessary for the receiver to be able to reconstruct the message, as well as
our prior beliefs of the model.
The message must include:

\begin{enumerate}
  \item The number of Gaussian mixtures, $K$.
  \item Encoding the relative weights $\weights$ of the $K$ Gaussian mixtures.
  \item Encoding the component parameters $\vecmean$, $\veccov$ for all $K$
        Gaussian mixtures.
  \item Encoding the data, given the model parameters.
  \item The lattice quantization constant $\kappa(q)$ for the number of model
        parameters $q$.
\end{enumerate}

Some of these components require non-trivial derivations or approximations in
order to calculate the length of the optimal loseless encoding. 
For this reason, we will consider the message length of each component in turn
before returning to the full message length specified in Equation \ref{eq:message-length}.

@@

We assume that fewer mixtures are more likely than a larger number of mixtures.
Specifically, we assume that $p(K) \propto 2^{-K}$, which implies that the
optimal message length required to encode $K$ is given by:

@@

We assume a uniform prior on the mixing weights $\bm{w}$, only requiring that
$\sum_{k=1}^{K}w_k = 1$. 
This implies that the weights can be treated as parameters of a multinomial
distribution, and the length of their optimally encoded message is given by
\citet{Boulton_1969},

\begin{equation}
  I(\weights) 
    = \frac{K - 1}{2}\log{N} 
    - \frac{1}{2}\sum_{k=1}^{K}\log\weight_k 
    - \log{\Gamma{\left(K\right)}} \quad ,
\end{equation}

@@

In order to properly encode the mixture parameters $\vecmean$ and $\veccov$
for all $K$ mixtures, we must encode both our prior belief on those parameters,
and the determinant of the expected Fisher information matrix.
For the $k$-th mixture this becomes,

@@

\begin{equation}
  I(\vecmean,\veccov) = -\sum_{k=1}^{K}\log{p(\vecmean_k,\veccov_k)} + \frac{1}{2}\sum_{k=1}^{K}\log{|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|}
  \label{eq:10}
\end{equation}

We introduce our priors on $\vecmean$ and $\veccov$ before approximating the
expected Fisher information matrix.
We adopt a uniform prior of $\mathcal{U}(\vecmean) = [-\infinity, \infinity]$ 
on all multivariate mean abundances $\vecmean$ (in all $D$ dimensions).
We note that while this prior is improper, it becomes proper when it enters
the (accepted) posterior distribution, and it has no impact on our inferences.
Indeed, we similarly could adopt a (proper and) large uniform prior on all 
mean abundances of $\mathcal{U}(\vecmean) = [-12, 12]$ (24 orders of magnitude!)
and our inference would be the same.

We adopt a congujate inverted Wishart prior for the covariance matrices in the
individual mixtures. 
The joint prior density for the parameters of a single
mixture is given by the limiting form of the normal inverted-Wishart
density \citep[e.g., Section 5.2.3 of ][]{Schafer_1997}:

\begin{equation}
    p(\vecmean_k,\veccov_k) \propto |\veccov_k|^{\frac{D + 1}{2}} \quad .
    \label{eq:prior-mean-cov}
\end{equation}

Computing the expected Fisher information $\mathcal{F}(\vecmean_k,\veccov_k)$ 
requires the second order partial derivatives of the negative log-likelihood 
$-\likelihood(\data|\vecmean,\veccov)$.
We approximate the determinant of the Fisher information 
$|\mathcal{F}(\vecmean_k, \veccov_k)|$ as the product of 
$|\mathcal{F}\left(\vecmean_k\right)|$ and $|\mathcal{F}\left(\veccov_k\right)|$ 
\citep{Oliver_1996,Roberts_1998}. 
We begin by taking second derivative of the log-likelihood function
$\likelihood\left(\data|\vecmean,\veccov\right)$,

@@

\noindent{}and for a $D$-dimensional vector $\vecmean$ the Fisher information
matrix is:

\begin{eqnarray}
  |\mathcal{F}(\vecmean_k)| = \left|-\frac{\delta^2\likelihood}{\delta\vecmean^2}\right| = N^D|\veccov|^{-1} \quad .
  \label{eq:approx-fisher-mean}
\end{eqnarray}

Computing the Fisher information of the covariance matrix 
$|\mathcal{F}\left(\veccov\right)|$ is harder.
Using theory of matrix derivatives based on matrix vectorization \citep{Dwyer_1967},
\citet{Magnus_1988} derived the following analytical expression for the 
determinant of the Fisher information for a covariance matrix
$|\mathcal{F}\left(\veccov_k\right)|$,

% TODO: This derivation is a nightmare to find

\begin{equation}
  |\mathcal{F}\left(\veccov_k\right)| = N^\frac{D(D+1)}{2}2^{-D}|\veccov_k|^{-(D+1)}
  \label{eq:approx-fisher-cov}
\end{equation}

\noindent{}where $D(D+1)/2$ arises from the number of free parameters in a
covariance matrix with off-diagonal terms.
Equations \ref{eq:approx-fisher-mean} and \ref{eq:approx-fisher-cov} allow us
to approximate the determinant of the expected Fisher information matrix
$|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|$ as:

\begin{eqnarray}
  |\mathcal{F}\left(\vecmean_k,\veccov_k\right)| & \approx & |\mathcal{F}\left(\vecmean_k\right)|\cdot|\mathcal{F}\left(\veccov_k\right)| \nonumber \\
  |\mathcal{F}\left(\vecmean_k,\veccov_k\right)| & \approx & N^{D}|\veccov_k|^{-1}N^\frac{D(D+1)}{2}2^{-D}|\veccov_k|^{-(D+1)} \nonumber \\
  |\mathcal{F}\left(\vecmean_k,\veccov_k\right)| & \approx & N^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)} \quad .
  \label{eq:fisher-mean-cov}
\end{eqnarray}

Thus for all $K$ mixtures, substituting Equations \ref{eq:prior-mean-cov} and 
\ref{eq:fisher-mean-cov} into Equation \ref{eq:10} gives:

\begin{eqnarray}
I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\log{p(\vecmean_k,\veccov_k)} + \frac{1}{2}\sum_{k=1}^{K}\log{|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|} \nonumber  \\
I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\log{\left(|\veccov_k|^{\frac{D + 1}{2}}\right)}
 + \frac{1}{2}\sum_{k=1}^{K}\log{\left(N^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)}\right)} \nonumber \\
I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\frac{D + 1}{2}\log|\veccov_k|
 + \frac{1}{2}\sum_{k=1}^{K}\left[\frac{D(D+3)}{2}\log{N} -D\log{2} -(D+2)\log|\veccov_k|\right] \nonumber \\
I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\frac{D + 1}{2}\log|\veccov_k|
 + \frac{1}{2}K\left[\frac{D\left(D+3\right)}{2}\log{N} - D\log{2}\right]
 -\sum_{k=1}^{K}\frac{D+2}{2}\log|\veccov_k| \nonumber \\
I(\vecmean,\veccov) &=& \frac{1}{2}\sum_{k=1}^{K}\log|\veccov_k|
 + \frac{1}{2}K\left[\frac{D\left(D+3\right)}{2}\log{N} - D\log{2}\right] \quad .
\end{eqnarray}

\subsubsection{Encoding the data, given the model parameters}
\label{sec:encoding-data}

Each datum can only be stated with a precision which is given by the accuracy
of the measurement.
If we assume homoskedastic noise properties for the observed data, and the
precision that each elemental abundance can be measured is $\mathcal{E}$, then
the \emph{probability} of a datum is given by 
$\textrm{Pr}(\datum_n) = \mathcal{E}^D\textrm{Pr}(\datum_i|\mathcal{M})$
where $\mathcal{M}$ represents the model and $\textrm{Pr}(\datum_i|\mathcal{M})$
is the \emph{probability density}.
Note that $\mathcal{E}$ is a constant, and is only used to maintain validity 
between \emph{probability} and \emph{probability density}.
For these experiments we adopt $\mathcal{E} = 0.01$, but note that the value of
$\mathcal{E}$ has no effect on our inferences.

The length of the total encoding of a datum given the model is,
\begin{equation}
  I(\data_n) = -\log{\textrm{Pr}(\data_n)} = -D\log\mathcal{E} - \log\sum_{k=1}^{K}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)
\end{equation}

\noindent{}and for the entire data:

\begin{eqnarray}
  I(\data|\vectheta) &=& -ND\log\mathcal{E} - \sum_{n=1}^{N}\log\sum_{k=1}^{K}w_{k}f_k(\data_n|\vecmean_k,\veccov_k) \quad .\\
%  I(\data|\vectheta) &=& -ND\log\mathcal{E} - \sum_{k=1}^{K}\log{\weight_k} - \sum_{n=1}^{N}\sum_{k=1}^{K}f_k\left(\data_n|\vecmean_k,\veccov_k\right) \quad .
\end{eqnarray}

\subsubsection{Encoding the lattice constant}
\label{sec:encoding-lattice-constant}

The lattice quantisation constant arises from the approximation to the (so-called)
strict MML, where parameters are quantised into intervals in high dimensional
space.
First we will calculate the total number of free parameters in our model.
Recall that we assume our covariance matrices have non-zero off-diagonal terms,
requiring $\frac{KD(D+1)}{2}$ parameters for all $K$ covariance matrices
(where we implicitly assume $D > 1$).
The $K$ $D$-dimensional vector means require a total of $KD$ parameters.
Under the assumption $K > 1$, the $K$ relative weights $\weights$ only require
$K - 1$ parameters because $\sum_{k=1}^{K}\weight_k = 1$.
Therefore the total number of free parameters $Q$ is given by,

\begin{eqnarray}
  Q &=& \frac{KD(D+1)}{2} + KD + K - 1 \nonumber \\
  Q &=& K\left[\frac{D\left(D+3\right)}{2} + 1\right] - 1 \quad .
  \label{eq:q}
\end{eqnarray}

We use an approximation for the logarithm of the lattice constant
\citep[see Sections 5.2.12 and 3.3.4 of ][]{Wallace_2005} where,

\begin{eqnarray}
    I(y) \approx - \log\frac{p(\theta')}{\sqrt{F(\theta')}} 
                   - \log{f(y|\theta')}
                   - (Q/2)\log{2\pi}
                   + \frac{1}{2}\log{(\pi{}Q)}
                   - 1
\end{eqnarray}

\noindent{}such that,
\begin{eqnarray}
  \log\kappa(Q) &=& \frac{\log{Q\pi}}{Q} - \log{2\pi} - 1
\end{eqnarray}

\noindent{}and the expected error on $\log\kappa(Q)$ is less than 0.1~nit.
The $Q$ (lattice) terms in Equation \ref{eq:message-length} becomes:
